\documentclass[11pt]{article}
\usepackage{times}
\usepackage{fullpage}
\usepackage{amsmath,amssymb}
\usepackage{graphicx}
\usepackage[colorlinks=true,linkcolor=blue,citecolor=blue,urlcolor=blue]{hyperref}
\usepackage{booktabs}
\usepackage{enumitem}
\usepackage{xcolor}
\usepackage{microtype}
\usepackage{fancyhdr}

\definecolor{accent}{HTML}{3C3489}
\definecolor{teal}{HTML}{085041}

\pagestyle{fancy}
\fancyhf{}
\rhead{\textcolor{accent}{Mythos Standard v1.0}}
\lhead{Attractor Dynamics Research Lab}
\cfoot{\thepage}

\title{\textbf{Mythos Standard v1.0}\\
\large A Unified Framework for Governed AI Systems}
\author{Zetta Byte\\
Attractor Dynamics Research Lab\\
\texttt{10.5281/zenodo.19464889}}
\date{April 2026}

\begin{document}
\maketitle
\thispagestyle{fancy}

\begin{abstract}
The Mythos Standard defines a unified framework for governed AI systems spanning mathematical
foundations, governance substrates, runtime monitoring, threat modeling, and containment
enforcement. It specifies the interfaces, invariants, schemas, and evaluation procedures required
to implement a Mythos-Class governed runtime. The standard ensures reproducibility,
interoperability, formal governance guarantees, and consistent evaluation across institutions.
\end{abstract}

\tableofcontents
\newpage

%% ── PREAMBLE ─────────────────────────────────────────────────────────────────

\section{Preamble}

The Mythos Standard is organized into seven layers:

\begin{enumerate}[noitemsep]
\item \textbf{Mathematical Foundations} --- exact curvature engines (hcderiv, curvopt)
\item \textbf{Governance Substrate} --- Constitutional OS
\item \textbf{Runtime Monitoring} --- CARE drift and deviation detection
\item \textbf{Threat Modeling Layer} --- machine-readable registry and simulator scenarios
\item \textbf{Containment Architecture} --- 8-layer enforcement model with membrane consensus
\item \textbf{Applied Runtime} --- governed multi-agent execution
\item \textbf{Evaluation Protocol} --- sim scenarios, harness probes, registry updates
\end{enumerate}

The standard ensures: reproducibility, interoperability, formal governance guarantees,
consistent evaluation across institutions, and compatibility with future Mythos-Class runtimes.

%% ── CONCEPTUAL FOUNDATIONS ───────────────────────────────────────────────────

\section{Conceptual Foundations}

\subsection{Unified Attractor Geometry (UAG)}

Any governed system is modelled as a trajectory in a high-dimensional state space partitioned
into attractor basins separated by ridges. Three UAG primitives are operationalized across the
stack: $\kappa$ (curvature), $\delta$ (drift), and basin transitions. Transition probability is
governed by the Kramers escape rate~\cite{byte2026uag}:

\begin{equation}
S(\gamma^*) \approx \frac{2\,\Delta R}{|\lambda_{\min}(H(\mathbf{x}^*))|}
\end{equation}

where $\Delta R$ is the energy gap to the ridge crest and $\lambda_{\min}$ is the smallest
Hessian eigenvalue at the saddle point $\mathbf{x}^*$.

\subsection{Lyapunov Stability}

Governance stability is certified by:

\begin{equation}
V(t) = \alpha \cdot c(t) + \beta \cdot u(t) + \gamma \cdot d(t)
\end{equation}

where $c(t)$ is a curvature-derived complexity term, $u(t)$ an uncertainty term, and $d(t)$ a
drift term. Stability requires $V(t+1) \leq V(t)$ for all $t$ under all admitted deltas.

\subsection{Membrane Theory}

A membrane $m$ enforces a predicate $P_m$ over proposed deltas. A delta $\delta$ is
\textbf{admitted} by membrane $m$ iff $P_m(\delta, s) = \mathrm{true}$. Five membranes
(M1--M5) vote in parallel; 3-of-5 majority required. M5 (curvature-drift) carries veto power.

%% ── FORMAL DEFINITIONS ───────────────────────────────────────────────────────

\section{Formal Definitions}

\paragraph{Governance profile.}
A governance profile is a typed configuration $G$ specifying active rules, membranes, and
invariants:
\[
G = \langle R, M, I \rangle
\]
where $R$ is a set of rules, $M$ a set of membranes, and $I$ a set of invariants.

\paragraph{Delta.}
A delta $\delta$ is a typed, reversible state transformation on governed system state $s$:
\[
\delta : S \to S, \quad \exists\,\delta^{-1} : S \to S \text{ such that } \delta^{-1}(\delta(s)) = s
\]
Deltas compose associatively and must preserve all active invariants in $I$.

\paragraph{Membrane.}
A membrane $m$ is a guarded interface enforcing predicate $P_m$ over proposed deltas.
$\delta$ is admitted by $m$ iff $P_m(\delta, s) = \mathrm{true}$.

\paragraph{Invariant.}
An invariant $i \in I$ is a predicate over system states and histories:
\[
i : S \times H \to \{\mathrm{true}, \mathrm{false}\}
\]
A runtime is compliant iff all $i \in I$ remain true for all executions.

\paragraph{Human primacy invariant ($i_{\mathrm{HP}}$).}
Any unresolved or high-risk decision path must be escalated to a human operator. No autonomous
path may bypass this escalation channel.

\paragraph{Drift.}
Drift is a measure of distributional or behavioral deviation between reference behavior
$B_{\mathrm{ref}}$ and current behavior $B_t$, typically instantiated as a divergence
$D(B_t \| B_{\mathrm{ref}})$ (e.g., KL over TF-IDF or embedding distributions).

\paragraph{Threat.}
A threat $T$ is a structured record in the registry:
\[
T = \langle \mathrm{id}, \mathrm{class}, \mathrm{description}, \mathrm{scenarios},
            \mathrm{metrics}, \mathrm{coverage\_status} \rangle
\]
with $\mathrm{coverage\_status} \in \{\mathrm{none}, \mathrm{partial}, \mathrm{full}\}$.

\paragraph{Simulator scenario.}
A simulator scenario $S$ is a parameterized evaluation setting:
\[
S = \langle \mathrm{id}, \mathrm{prompt\_family}, \mathrm{model\_config},
            \mathrm{runtime\_config} \rangle
\]

\paragraph{Harness specification (HAR-SPEC).}
A harness spec $H$ defines: required inputs, required metrics, pass/fail conditions, and
reporting format. A system satisfies $H$ iff it can execute the specified runs and produce
metrics meeting the defined thresholds.

%% ── MATHEMATICAL FOUNDATIONS ─────────────────────────────────────────────────

\section{Mathematical Foundations}

\subsection{hcderiv --- Exact One-Pass Hessians}

\texttt{hcderiv} computes exact second-order derivatives via hypercomplex perturbation algebra
in a single forward pass. The JAX-XLA backend represents the full coefficient vector as a
single JAX array implementing the Hyper algebra via \texttt{jnp.einsum}~\cite{byte2026hcderiv}.

\noindent \textbf{Required interface:}
\texttt{hessian(f, x, backend)} returning $H \in \mathbb{R}^{n \times n}$.

\subsection{curvopt --- Curvature-Aware Optimization}

\texttt{curvopt} uses exact Hessians from \texttt{hcderiv} to set trust-region radius from
ridge curvature (UAG Theorem~4). Required interface: \texttt{optimizer(f, x0, hessian\_fn)}
returning $(x^*, \mathrm{trace})$.

%% ── GOVERNANCE SUBSTRATE ─────────────────────────────────────────────────────

\section{Governance Substrate}

Constitutional OS formalizes governance as:
\[
\Sigma = (\Sigma_R, \Sigma_C, \Sigma_X), \quad \Phi = G \circ E
\]

\noindent \textbf{Governance invariants} (all Mythos-compliant systems must enforce):
\begin{itemize}[noitemsep]
\item \textbf{Continuity} --- no state transition may sever the audit chain
\item \textbf{Non-contradiction} --- no two admitted deltas may produce contradictory states
\item \textbf{Human primacy} ($i_{\mathrm{HP}}$) --- all high-risk paths escalate to human
\item \textbf{Escalation correctness} --- all detected deviations above threshold are routed
      to the correct oversight layer
\end{itemize}

%% ── RUNTIME MONITORING ───────────────────────────────────────────────────────

\section{Runtime Monitoring}

CARE computes behavioral deviation fingerprint $F_s = (\kappa, \beta, \tau, \delta_{\mathrm{UAG}})$
and evaluates risk potential $R(\mathbf{x})$ via exact Hessian from \texttt{hcderiv}. M5 fires
when risk delta exceeds 25\% of baseline, triggering freeze regardless of other membrane votes.

\noindent \textbf{Runtime actions:} Freeze, Rollback, Sentinel mode (unilateral OOBS control).

\noindent \textbf{Logging requirements:} Append-only drift log, deviation log,
rule-activation log; external storage required (Redis Streams or S3 with write-only credentials).

%% ── THREAT MODELING ──────────────────────────────────────────────────────────

\section{Threat Modeling Layer}

\subsection{Registry Schema}

Each threat entry $T$ is a JSON object with required fields:
\texttt{threat\_id} (TM-NNN), \texttt{threat\_class}, \texttt{severity}
$\in \{$\texttt{low}, \texttt{medium}, \texttt{high}, \texttt{catastrophic}$\}$,
\texttt{exploitability} $\in \{$\texttt{opportunistic}, \texttt{skilled},
\texttt{frontier\_only}$\}$, \texttt{coverage\_status} $\in \{$\texttt{none},
\texttt{partial}, \texttt{full}$\}$, \texttt{tested\_by\_sim}, \texttt{tested\_by\_harness}.

\subsection{Threat Classes (v0.1)}

\begin{table}[h]
\centering
\begin{tabular}{lll}
\toprule
ID & Class & Severity \\
\midrule
TM-002 & governance\_drift & high \\
TM-005 & cross\_model\_agreement\_failure & high \\
TM-012 & latent\_tool\_discovery & catastrophic \\
TM-009 & specification\_gaming (v1.1) & high \\
TM-015 & governance\_deadlock (v1.1) & high \\
\bottomrule
\end{tabular}
\end{table}

%% ── CONTAINMENT ARCHITECTURE ─────────────────────────────────────────────────

\section{Containment Architecture}

\begin{table}[h]
\centering
\begin{tabular}{cll}
\toprule
Layer & Name & Mechanism \\
\midrule
1 & I/O Membrane & Schema validation, depth limits \\
2 & Outbound Strangulation & Default-deny egress, HMAC calls \\
3 & Behavioral Fingerprinting & Token curvature, plan topology \\
4 & Multi-Membrane Consensus & 3-of-5 vote across M1--M5 \\
5 & Out-of-Band Sentinel & Separate machine, freeze/rollback \\
6 & Deterministic Replay & Append-only audit log \\
7 & Cold-Storage Zone & Air-gapped GPU jail \\
8 & Cross-Model Verification & Two-pilot cockpit \\
\bottomrule
\end{tabular}
\end{table}

\noindent \textbf{M5 (canary-drift membrane)} carries veto power: CARE risk delta $>$ 25\%
triggers immediate freeze independent of the 3-of-5 vote.

%% ── EVALUATION PROTOCOL ──────────────────────────────────────────────────────

\section{Evaluation Protocol}

\subsection{Key Metrics}

\begin{table}[h]
\centering
\begin{tabular}{lll}
\toprule
Metric & Definition & Formula \\
\midrule
RAC & Rule-Activation Consistency &
  $\frac{\#\{\text{responses with governance signal}\}}{\#\{\text{responses}\}}$ \\[4pt]
DSI & Drift Stability Index &
  $D(F \,\|\, F_{\mathrm{base}})$ (normalised) \\[4pt]
ADR & Adversarial Detection Rate &
  $\frac{\#\{\text{true divergences detected}\}}{\#\{\text{true divergences}\}}$ \\[4pt]
FNR & False-Negative Rate &
  $\frac{\#\{\text{missed divergences}\}}{\#\{\text{true divergences}\}}$ \\[4pt]
FPR & False-Positive Rate &
  $\frac{\#\{\text{non-divergent cases flagged}\}}{\#\{\text{non-divergent cases}\}}$ \\[4pt]
MC  & Mitigation Correctness &
  $\frac{\#\{\text{correct mitigations}\}}{\#\{\text{detected divergences}\}}$ \\[4pt]
NDR & Novelty Detection Rate &
  $\frac{\#\{\text{novel behaviors detected}\}}{\#\{\text{novel behaviors}\}}$ \\[4pt]
SER & Silent Expansion Rate &
  $\frac{\#\{\text{novel behaviors silent}\}}{\#\{\text{novel behaviors}\}}$ \\[4pt]
EC  & Escalation Correctness &
  $\frac{\#\{\text{deviations escalated correctly}\}}{\#\{\text{deviations}\}}$ \\[4pt]
BF  & Baseline Fit &
  $\mathrm{sim}(B_{\mathrm{obs}}, B_{\mathrm{ref}})$ \\
\bottomrule
\end{tabular}
\end{table}

\noindent \textbf{Hard gate:} SER $\leq 0.05$ for all compliant evaluations.

\subsection{Threshold Calibration}

All thresholds (DSI gate, FPR gate, BF gate, EC gate) must be set from pilot data before
full evaluation runs. Document all choices in \texttt{section7\_numbers\_calibrated.json}.

%% ── COMPLIANCE ───────────────────────────────────────────────────────────────

\section{Compliance}

\begin{table}[h]
\centering
\begin{tabular}{ll}
\toprule
Tier & Requirements \\
\midrule
Registry-only & L5 only --- consumable threat registry, no runtime enforcement \\
3-layer starter & L3 + L4 + CMV --- minimum viable containment posture \\
Full v0.1 & All seven layers + v0.1 harness passing \\
Full v1.1 & All layers + expanded harness + public trace corpus \\
\bottomrule
\end{tabular}
\end{table}

%% ── REFERENCE IMPLEMENTATION ─────────────────────────────────────────────────

\section{Reference Implementation}

A Mythos-compliant reference implementation must demonstrate the full evaluation loop:
load COS governance profile $\to$ start CARE monitor $\to$ load threat registry $\to$
execute one SIM scenario $\to$ run one HAR-SPEC probe $\to$ produce registry update.

\noindent \textbf{docker-compose}: \texttt{docker-compose up mythos-demo}

\noindent \textbf{Starter kit} must include: \texttt{registry/}, \texttt{sims/},
\texttt{harness/}, \texttt{cos\_ruleset/}, \texttt{care\_config/}, \texttt{README.md}.

%% ── REFERENCES ───────────────────────────────────────────────────────────────

\bibliographystyle{plain}
\begin{thebibliography}{9}

\bibitem{byte2026uag}
Byte, Z. \textit{The Unified Attractor Grammar}. Zenodo (2026).
DOI: \href{https://doi.org/10.5281/zenodo.19448508}{10.5281/zenodo.19448508}

\bibitem{byte2026cos}
Byte, Z. \textit{Constitutional OS: A Formal Governance Substrate for AI Ecosystems}.
Zenodo (2026). DOI: \href{https://doi.org/10.5281/zenodo.19258310}{10.5281/zenodo.19258310}

\bibitem{byte2026mcca}
Byte, Z. \textit{Mythos-Class Containment Architecture}. Zenodo (2026).
DOI: \href{https://doi.org/10.5281/zenodo.19464889}{10.5281/zenodo.19464889}

\bibitem{byte2026hcderiv}
Byte, Z. \textit{One-Pass Exact Hessians via Hypercomplex Perturbation} (hcderiv v4.1).
Zenodo (2026). DOI: \href{https://doi.org/10.5281/zenodo.19476000}{10.5281/zenodo.19476000}

\end{thebibliography}

\end{document}
